Supplementary Materials:\\Black-swan events in animal populations

### 2016-05-24 text/ms.tex
Copyediting of main paper

\textbf{Abstract}:
Black swans are improbable events that nonetheless occur---often with profound consequences. While physical extremes, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency, it remains unclear the extent to which ecological population numbers buffer or suffer from such extremes. Here we estimate the frequency and direction of black-swan events in 609 animal populations, finding strong evidence for their occurrence in \(\sim\)\overallBasePerc \% of populations. Black-swan events occur most frequently for birds (\birdPH \%), mammals (\mammalsPH \%), and insects (\insectsPH \%), and manifest primarily as   population die-offs and crashes (\percBSDown \%) driven by extrinsic perturbations from climate variability. Thus, ignoring black-swan events will under-estimate extinction risk. Our results demonstrate the importance of both modeling heavy-tailed downward events in populations, and developing conservation strategies that are robust to future increases in climate extremes.

\textbf{One Sentence Summary}: Black swans---statistically improbable events with profound consequences---occur in animal populations and understanding them will improve climate risk assessments.

@@

\textbf{Main text}:
Major surprises happen more often than expected in financial, social, and natural systems\cite{taleb2007, sornette2009, may2008}. Massive unpredictable market swings are responsible for the majority of financial gains and losses\cite{taleb2007}, fatalities from the largest wars dwarf those from all others \cite{newman2005}, and the frequency of the most damaging earthquakes has exceeded past expectations\cite{sornette2009}. In ecological systems, background rates of global extinction are punctuated by mass extinction\cite{harnik2012}, viruses can mutate suddenly to infect new hosts, and billions of animals can die at once in mass mortality events\cite{fey2015}. Indeed, such die-offs may be the most important element affecting population persistence\cite{mangel1994} and their importance is likely to increase given projected increases in the frequency and magnitude of climate-related extremes\cite{ipcc2012}. Despite this anecdotal evidence for ecological black swans, systemtic searches of large numbers of timeseries have yet to reveal black-swan events\cite{keitt1998, allen2001, halley2002} However, the overwhelming majority of model fitting and risk forecasting assumes that deviations from model predictions can be represented by a normal distribution\cite[e.g.][]{brook2006a, knape2012}. If black swans occur, though, a normal distribution would under-estimate the probability of extreme events occurring.

@@

Specifically, we fit population dynamics models in which the process noise is drawn from a Student-t distribution. By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are therefore diagnositc of black-swan events (Fig.~1, fig.~S\ref{fig:priors}). Lower values of \(\nu\) result in heavy-tailed distributions. For example, at \(\nu = 2\), the probability of drawing a value more than five standard deviations below the mean is almost 70,000 times higher than the probability of drawing such a value from a normal distribution (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)). As \(\nu\) approaches infinity, the distribution approaches the normal distribution (Fig.~1).

We found evidence that black-swan dynamics are infreqent but highly influentiual: they occur most frequently for birds (7\%), mammals (5\%), and insects (3\%) and almost never in fishes for this dataset (Fig.~2). Black-swan events were taxonomically widespread, occurring in 38\% of taxonomic orders. Accounting for time series length and partially pooling inference across taxonomic class and order with a hierarchical model, we found stronger evidence for black swans in insect populations than these statistics suggest---four of eight orders with the highest median probability of heavy tails were insect orders (Fig.~3a).

@@

Given the prevalence of downwards events, we refit our heavy-tailed models to measure the degree of skewness \(\gamma\) of the process noise using a skew-t distribution (Fig.~4a), and used these models to make near-term risk forecasts. Aggregated across populations with strong evidence of heavy tails (median \(\nu < 10\)), 85\% of the \(\gamma\) probability density was below 1, indicating strong evidence of downwardly skewed process noise (Fig.~4b, fig.~S\ref{fig:skew-nu}). In contrast, populations that did not have heavy tails (median \(\nu \geq 70\)) had little evidence of skewed process noise (Fig.~4b; 90\% of 95\% credible intervals overlapped 1). Projecting these heavy-tailed populations forward five years revealed that assuming the standard normal process noise underestimated risk by \crashUnderRange-fold (interquartile range of 99\% lower credible interval of abundance; Fig.~4c, fig.~S\ref{fig:skew-projections}). Thus our analysis provides strong evidence for downward-skewed heavy-tailed events in abundance time series of higher taxa, and ignoring these events will tend to underestimate the risk of population declines.

% TODO How are you defining risk. See if you can make this more concrete to avoid IUCN people getting upset with lack of specificity (i.e. Resit akcakaya)

@@

In light of our findings, we suggest that ecological resource management can learn from disciplines that focus on heavy tails. For example, earthquake preparedness and response is focused on black-swan events. Similarly to ecological black swans, we can rarely predict the specific timing of large earthquakes. But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans\cite{nrc2007}. The presence of black swans also suggests that we develop management policy that is robust to heavy tails. For instance, setting target population abundances far from critical limits will buffer against black-swan events\cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations to persist when others are affected by disease or extreme environmental forces\cite{schindler2010}. Finally, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when models break down, and adjust future management policy\cite{doak2008, lindenmayer2010}.

Rare catastrophes can have a profound influence on population persistence\cite{mangel1994}. In recent decades, ecology has moved toward focusing on aspects of variance in addition to mean responses\cite{loreau2010a, thompson2013}. Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes\cite{meehl2004, ipcc2012, thompson2013}. Financial analysts are concerned with the shape of financial return downward tails because these directly impact estimates of risk---the probability of a specific magnitude of undesired event occurring\cite{rachev2008}. Similarly, ecologists should focus more on estimating and predicting downward tails of population abundance, since these increase true extinction risk.

@@

\textbf{Supplementary Materials}\\
Materials and Methods\\
References (32 to 72)\\
Tables S1 to S2\\
Figs.~S1 to S8\\

@@

% TODO The A, B, C are too close to the text in the figure.
\textbf{Fig.~4: Heavy-tailed process noise tends to be downward skewed and ignoring this can underestimate risk.} (\textbf{A}) Illustrations of Student-t distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)). (\textbf{B}) Posterior density of the skewness parameters aggregated across populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations (\(\hat{\nu} > 70\)). Approximate mid-values from \textbf{B} are illustrated in \textbf{A}. (\textbf{C}) Example time series of heavy-tailed populations with different levels of skewness. Red dots highlight likely heavy-tailed events. Forecasts (grey regions) show median (solid lines) and lower 99\% credible intervals (dotted lines) of abundance. Black and red lines indicate forecasts from Gompertz models with lognormal and skew-t process noise, respectively.

>>>>

\textbf{Abstract}: Black swans are improbable events that nonetheless
occur---often with profound consequences. While physical extremes, such as
monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological population numbers
buffer or suffer from such extremes. Here we estimate the frequency and
direction of black-swan events in \NPops\ animal populations, finding strong
evidence for their occurrence in \(\sim\)\overallBasePerc \% of populations.
Black-swan events occur most frequently for birds (\birdPH \%), mammals
(\mammalsPH \%), and insects (\insectsPH \%), and manifest primarily as
population die-offs and crashes (\percBSDown \%) driven by extrinsic
perturbations from climate variability. Thus, ignoring black-swan events will
underestimate extinction risk. Our results demonstrate the importance of both
modeling heavy-tailed downward events in populations, and developing
conservation strategies that are robust to future increases in climate
extremes.

% \textbf{One Sentence Summary}: Black swans---statistically improbable events with profound consequences---occur in animal populations and understanding them will improve climate risk assessments.

@@

\textbf{Main text}:
Major surprises happen more often than expected in financial, social, and natural systems\cite{taleb2007, sornette2009, may2008}. Massive unpredictable market swings are responsible for the majority of financial gains and losses\cite{taleb2007}, fatalities from the largest wars dwarf those from all others \cite{newman2005}, and the frequency of the most damaging earthquakes has exceeded past expectations\cite{sornette2009}. In ecological systems, background rates of global extinction are punctuated by mass extinction\cite{harnik2012}, viruses can mutate suddenly to infect new hosts, and billions of animals can die at once in mass mortality events\cite{fey2015}. Indeed, such die-offs may be the most important element affecting population persistence\cite{mangel1994} and their importance is likely to increase given projected increases in the frequency and magnitude of climate-related extremes\cite{ipcc2012}. Despite this anecdotal evidence for ecological black swans, systematic searches of large numbers of timeseries have yet to reveal black-swan events\cite{keitt1998, allen2001, halley2002} However, the overwhelming majority of model fitting and risk forecasting assumes that deviations from model predictions can be represented by a normal distribution\cite{brook2006a, knape2012}. If black swans occur, though, a normal distribution would under-estimate the probability of extreme events occurring.

@@

Specifically, we fit population dynamics models in which the process noise is drawn from a Student-t distribution. By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are therefore diagnostic of black-swan events (Fig.~1, fig.~S\ref{fig:priors}). Lower values of \(\nu\) result in heavy-tailed distributions. For example, at \(\nu = 2\), the probability of drawing a value more than five standard deviations below the mean is almost 70,000 times higher than the probability of drawing such a value from a normal distribution (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)). As \(\nu\) approaches infinity, the distribution approaches the normal distribution (Fig.~1).

We found evidence that black-swan dynamics are infrequent but highly influential: they occur most frequently for birds (7\%), mammals (5\%), and insects (3\%) and almost never in fishes for this dataset (Fig.~2). Black-swan events were taxonomically widespread, occurring in 38\% of taxonomic orders. Accounting for time series length and partially pooling inference across taxonomic class and order with a hierarchical model, we found stronger evidence for black swans in insect populations than these statistics suggest---four of eight orders with the highest median probability of heavy tails were insect orders (Fig.~3a).

@@

Given the prevalence of downwards events, we refit our heavy-tailed models to measure the degree of skewness \(\gamma\) of the process noise using a skew-t distribution (Fig.~4a), and used these models to make near-term risk forecasts. Aggregated across populations with strong evidence of heavy tails (median \(\nu < 10\)), 85\% of the \(\gamma\) probability density was below 1, indicating strong evidence of downwardly skewed process noise (Fig.~4b, fig.~S\ref{fig:skew-nu}). In contrast, populations that did not have heavy tails (median \(\nu \geq 70\)) had little evidence of skewed process noise (Fig.~4b; 90\% of 95\% credible intervals overlapped 1). Projecting these heavy-tailed populations forward five years revealed that assuming standard normal process noise underestimated risk (99\% lower credible interval of abundance) by \crashUnderRange-fold (interquartile range; Fig.~4c, fig.~S\ref{fig:skew-projections}). Thus our analysis provides strong evidence for downward-skewed heavy-tailed events in abundance time series of higher taxa, and ignoring these events will tend to underestimate the risk of population declines.

@@

In light of our findings, we suggest that ecological resource management can learn from disciplines that focus on heavy tails. For example, earthquake preparedness and response is focused on black-swan events. Similarly to ecological black swans, we can rarely predict the specific timing of large earthquakes. But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans\cite{nrc2007}. The presence of black swans also suggests that we develop management policy that is robust to heavy tails and encourages \textit{general} resilience\cite{carpenter2012}. For instance, setting target population abundances far from critical limits will buffer against black-swan events\cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations to persist when others are affected by disease or extreme environmental forces\cite{schindler2010}. Finally, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when models break down, and adjust future management policy\cite{doak2008, lindenmayer2010}.

Rare catastrophes can have a profound influence on population persistence\cite{mangel1994}. In recent decades, ecology has moved toward focusing on aspects of variance in addition to mean responses\cite{thompson2013}. Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes\cite{meehl2004, ipcc2012, thompson2013}. Financial analysts are concerned with the shape of financial return downward tails because these directly impact estimates of risk---the probability of a specific magnitude of undesired event occurring. Similarly, ecologists should focus more on estimating and predicting downward tails of population abundance, since these increase true extinction risk.

@@

\textbf{Supplementary Materials}\\
Materials and Methods\\
Supplementary References\\
Tables S1 to S2\\
Figs.~S1 to S8\\

@@

\textbf{Fig.~4: Heavy-tailed process noise tends to be downward skewed and ignoring this can underestimate risk.} (\textbf{A}) Illustrations of Student-t distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)). (\textbf{B}) Posterior density of the skewness parameters aggregated across populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations (\(\hat{\nu} > 70\)). Approximate mid-values from \textbf{B} are illustrated in \textbf{A}. (\textbf{C}) Example time series of heavy-tailed populations with different levels of skewness. Red dots highlight likely heavy-tailed events. Forecasts (grey regions) show median (solid lines) and lower 99\% credible intervals (dotted lines) of abundance. Black and red lines indicate forecasts from Gompertz models with lognormal and skew-t process noise, respectively.
